Lies, versus the lying liars who tell them: comment on Garbarino, Slonim and Villeval (2018)

### 2018-11-01 GSV-comment-brief.Rmd

>  Garbarino, Slonim and Villeval (2018) describe a new method to calculate the probability distribution 
>  of the number of liars in "coin flip" style experiments. I run simulations showing that their method almost always gives 
>  erroneous confidence intervals. I explain why and describe two better
>  ways to calculate probability distributions.

@Garbarino2018 -- GSV from here on -- introduce a method to estimate the prevalence of lying in coin flip style experiments,
where participants are paid for reporting a high outcome of a random variable, such as heads on a coin flip, 
and where the true coin outcome is not observed. The input of the method is the number of "heads" 
reports, $R$, from a sample of size $N$, along with the probability $P$ of the *low* outcome on
the random variable. They claim that their method can estimate the full distribution of
lying outcomes, and they recommend using it for confidence intervals, hypothesis testing and power 
calculations.

Table 3 of GSV reports Monte Carlo simulations, which show that the probability of a Type II error
can be up to 50 per cent. GSV define a Type II error as the 95% confidence interval not containing
the true value. A 95% confidence interval which only contains the true value 50% of the time
is unusual.

@@

nreps <- 1000L
rewrite_csv <- FALSE # set to TRUE to (re)create the simulations CSV file
cache_results <- TRUE # FALSE to recalculate all results. TRUE for speed.
estimate_lies_not_liars <- FALSE # if FALSE, use Liars/N as ground truth
                                # if TRUE, use (R-heads)/(N-heads)
```

```{r parameters}

params <-   list(
  N     = c(10, 50, 100, 500),
  Liars = seq(0, 1, 0.1),
  P     = c(.2, .5, .8)
)

ci_levels <- c(90L, 95L, 99L)
```

>>>>

>  Garbarino, Slonim and Villeval (2018) describe a new method to calculate the probability distribution 
>  of the number of liars in "coin flip" style experiments. However, their method appears 
>  to give erroneous confidence intervals. I explain why and describe two better
>  ways to calculate probability distributions.

Some people are honest, while others will lie whenever it benefits them. We would like to know the
prevalence of lying, since a society where many people lie is likely to experience
economic and social problems. Since we cannot simply ask people if they are liars, one way to
estimate the proportion of liars in a group is to ask them to report the result of a coin flip or
other random device, offering them a payment if they report heads. Liars don't always lie: they 
only lie when it benefits them. So they always report heads irrespective of the true coin flip.
^[@Garbarino2018 maintain this assumption and so shall I.] If there are many more heads than
we would expect by chance, we can assume many people are lying. But how many?

A naïve estimate would be that if e.g. 80 people of 100 report heads, then on average 50 really saw
heads and 60% (30/50) of the remainder are lying. More generally, from $R$ reports of a good outcome
in a sample of size $N$, where the _bad_ outcome happens with probability $P$, we can estimate that
a proportion $\frac{R/N - (1-P)}{P}$ are lying [@abeler2016preferences]. The problem with this approach is that the number of heads is not
fixed. If we see 1 out of 3 people reporting heads, this method estimates there are less than zero
liars. But it is still possible that everyone saw heads and 1 person lied.

@Garbarino2018 -- GSV from here on -- introduce an alternative method. They claim that their method 
corrects for this problem and can estimate the full distribution of
lying outcomes, and they recommend using it for confidence intervals, hypothesis testing and power 
calculations.

Table 3 of GSV reports Monte Carlo simulations, which show that the probability of a Type II error
can be up to 50 per cent. GSV define a Type II error as the 95% confidence interval not containing
the true value. A 95% confidence interval which only contains the true value 50% of the time
is unusual. What is going on?

@@

nreps <- 1000L
rewrite_csv <- FALSE # set to TRUE to (re)create the simulations CSV file
cache_results <- TRUE # FALSE to recalculate all results. TRUE for speed.
sample_from_pop <- TRUE # if TRUE, the number of liars in the sample is randomly drawn 
                        # from a population with `Liars` liars. 
                        # If FALSE it is fixed at `Liars * N`.
estimate_lies_not_liars <- FALSE # if FALSE, use Liars/N as ground truth
                                # if TRUE, use (R-heads)/(N-heads)
```
